---
title: "Analysis: Vancouver's *Mobi* Bike Share"
author: "Andrew Luyt"
date: "`r paste('<br>Last updated ',format(Sys.Date(), '%A %B %d, %Y'))`"
output: 
  github_document:
    toc: true
    toc_depth: 2
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages and load data, echo=TRUE, results="hide", message=FALSE}
library(tidyverse)
library(lubridate)
library(sf)         # Vancouver map
library(gganimate)
library(viridis)

source("mobi_functions.R")

RIDES = "data/mobi_rides.Rdata"
if(file.exists(RIDES)) {
  load(RIDES)
  df <- as_tibble(df) %>% 
    mutate(hour = hour(depart_time))
} else {
  stop("Please run clean_data.R to create the required data")
}

MAP = "data/vancouver_map.geojson"
if(file.exists(MAP)) {
  MAP <- st_read(MAP)
} else {
  stop("Please run clean_data.R to create the required data")
}

STANLEY_PARK = "data/parks-polygon-representation.geojson"
if(file.exists(STANLEY_PARK)) {
  STANLEY_PARK <- st_read(STANLEY_PARK)
  STANLEY_PARK <- STANLEY_PARK %>% filter(park_name == 'Stanley Park')
}

rm(RIDES)

# Helper function for vectors
angle_from_x_axis <- function(y,x) {
  angle <- atan2(y, x)
  if (y<0) {
    angle <- angle + 2 * pi
  }
  angle * 180 / pi
}

```

```{r subset df to june and july}
# df <- df %>% 
#   filter(month(depart_time) %in% c(6,7))
```

```{r determine which membership classes have been used recently}
df %>% 
  group_by(membership) %>% 
  summarise(most_recent_use = max(depart_time)) %>% 
  arrange(most_recent_use)
```

So the first four seem to be no longer used in the system.

```{r theming}
default_theme <- theme_set(theme_bw())
theme_update(panel.grid = element_blank(),
             legend.justification = "top",
             plot.title = element_text(size=14, face="bold"),
             axis.text.x = element_text(size = 10),
             plot.margin = unit(c(0.3,0,0.3,0), "cm"))

theme_map <- theme_bw() +
  theme(axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = 'white'),
        plot.title = element_text(size=20, face="bold"),
        legend.justification = "top",
        legend.position = "none", 
        plot.margin = unit(c(0.3,0,0.3,0), "cm"))

theme_map_dark <- theme_bw() +
  theme(axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = 'grey40'),
        plot.title = element_text(size=20, face="bold"),
        legend.justification = "top",
        legend.position = "none", 
        plot.margin = unit(c(0.3,0,0.3,0), "cm"))
```

```{r arrange membership factors by number of rides}
ordered_memberships <- df %>% 
  group_by(membership) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  pull(membership) %>% 
  as.vector()
df <- df %>% 
  mutate(membership = fct_relevel(membership, ordered_memberships))
```


```{r plot rides by membership}
df %>% 
  group_by(membership) %>% tally() %>%  ggplot(aes(membership, n)) + geom_col() + coord_flip()
```

There are 20 membership types listed, but on the [https://www.mobibikes.ca/en/offers-subscription](webpage),
only 6 are listed. The others seem to be grandfathered in, or special offers
that aren't generally available.  We'll limit our analysis to a subset of
the most popular memberships. Two of these popular memberships, the 
**Community Pass** and **365 Day Founding Plus** are not currently listed as available.

```{r}
popular_memberships <- df %>% 
  group_by(membership) %>% 
  tally() %>% 
  filter(n > 9000) %>% 
  arrange(desc(n)) %>% 
  pull(membership)
popular_memberships <- popular_memberships[1:8]
popular_memberships <- as.factor(as.character(popular_memberships))
```


```{r}
df %>% 
  filter(membership %in% popular_memberships[1:6]) %>% 
  group_by(day = as.Date(depart_time), membership) %>% 
  tally() %>% 
  ggplot(aes(x = day, y = n, color = membership)) +
  geom_line(alpha = 0.5) + geom_smooth(se=FALSE) +
  labs(title = "Daily rides")
```

The 30 day and 24 hour passes are the most popular. The 365 passes seem to
be bought by those who don't mind riding in winter, as their is much less
seasonal variation in rides for those members.

Particularly in summer there's a lot of daily variation in rides. 

### TODO 
Is it from rain or what? You could
get rainfall data (downtown vancouver, not the airport?) and plot/corr mm of rain
with n_rides.

```{r rides by weekday and membership}
df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(weekday = wday(depart_time, label = TRUE, week_start = 1), membership) %>% 
  ggplot(aes(weekday, fill = membership)) +
  geom_bar(position = 'stack', color = 'black')
```
Weekends are really popular for 24-hour passes. The 365 passes are **more**
popular during the **week**, suggesting they might be being used for commuting.

```{r rides by weekday and membership 365 only}
df %>% 
  filter(membership %in% popular_memberships,
         str_detect(membership, "365")) %>% 
  group_by(weekday = wday(depart_time, label = TRUE, week_start = 1), membership) %>% 
  ggplot(aes(weekday, fill = membership)) +
  geom_bar(position = 'stack', color = 'black')
```

```{r look for patterns in mean by hour}
x <- df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(hour, membership) %>% 
  summarize(across(where(is.numeric), mean))
x %>% ggplot(aes(hour, trip_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, trip_distance)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, n_stopovers)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, stopover_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
#x %>% ggplot(aes(hour, kph)) + geom_col() + facet_wrap(~membership, ncol = 4)
```

* 24 hour passes have the longest rides (in time and distance)
* The 365 passes (especially Founding Plus) have more stopovers per trip
  * and also longer stopovers on average

```{r look for patterns in mean by weekday}
x <- df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(weekday = wday(depart_time, label = TRUE, week_start = 1), membership) %>% 
  summarize(across(where(is.numeric), mean))
x %>% ggplot(aes(weekday, trip_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, trip_distance)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, n_stopovers)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, stopover_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, kph)) + geom_col() + facet_wrap(~membership, ncol = 4)
```

* Trip durations and distances are very steady throughout the week
* 24 hour pass holders ride the slowest.
  * We noted above they also ride the longest duration and most distance.


The **365 Day Founding Plus** pass is a real outlier. It has many times more
stopovers per ride than any other type of pass.

### Stopovers?
Mobi bikes offer a [https://www.mobibikes.ca/en/news/make-quick-stop-stopover-mode](stopover mode) 
where the bike's steering column locks and
the integrated lock can be accessed. It is meant as a convenience when you 
need to stop somewhere that doesn't have a Mobi docking station, like a grocery
store. The ride timer doesn't stop and most Mobi memberships offer a maximum
of 30 minutes of use before overage fees start, so it would be interesting
to see what kinds of members have these stopovers on their trips. Let's
graph rides by the number of stopovers taken then colour by the membership type.

```{r number of stopovers by membership totals}
df %>% 
  filter(membership %in% popular_memberships) %>%
  # filter(membership == "365 Day Founding Plus") %>% 
  group_by(membership, n_stopovers) %>%
  ggplot(aes(x = n_stopovers, fill = membership)) +
  geom_bar()
```

As a first observation we see clearly that most people are taking a bike,
riding immediately to a desination and then docking the bike. Only a comparative
few stop somewhere along the way. Let's examine *only* those trips with at least
1 stopover.

```{r number of stopovers by membership totals only with stopovers}
df %>% 
  filter(membership %in% popular_memberships,
         n_stopovers > 0) %>%
  # filter(membership == "365 Day Founding Plus") %>% 
  group_by(membership, n_stopovers) %>%
  ggplot(aes(x = n_stopovers, fill = membership)) +
  geom_bar() + 
  labs(title = "Trips with at least 1 stopover") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 15))
```

Rides with one stopover have a broad distribution of memberships.
How about 2 or more stopovers?

```{r number of stopovers by membership totals only with stopovers}
df %>% 
  filter(membership %in% popular_memberships,
         n_stopovers > 1) %>%
  # filter(membership == "365 Day Founding Plus") %>% 
  group_by(membership, n_stopovers) %>%
  ggplot(aes(x = n_stopovers, fill = membership)) +
  geom_bar() + 
  labs(title = "Trips with at least 2 stopovers") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 15))
```

It looks like **365 Plus** and **365 Founding Plus** are the types of
memberships used for trips with a few stopovers.  Let's draw a graph
of *proportions* of rides, rather than total number of rides.

```{r number of stopovers by membership proportions}
df %>% 
  filter(membership %in% popular_memberships) %>%
  # filter(membership == "365 Day Founding Plus") %>% 
  group_by(membership, n_stopovers) %>%
  ggplot(aes(x = n_stopovers, fill = membership)) +
  geom_bar(position = "fill") +
  labs(title = "Which members are taking a stopover during a ride?",
       subtitle = "As the number of stopovers increases, it becomes more likely the rider has a\n365 Day Founding Plus Pass",
       y = "", x = "Stopovers") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
  scale_y_continuous(labels = scales::percent)
```

**365 Day Founding Plus** members make up most of the users that stop at a location during their
ride. **365 Day Plus** is in second place.  These both offer a notable clause in
their [https://www.mobibikes.ca/en/terms-and-conditions-use](agreements.) 
Most memberships offer 30 minutes of riding, after which
overage fees begin, but these two passes offer **60 minutes**. If a person
intends to use the bike system for running errands, going somewhere for
lunch, or other purposes which require more time than a simple A to B trip, 
these passes offer the best value by avoiding overage fees.

On the [https://www.mobibikes.ca/en/offers-subscription](offers page) however
we don't see the **365 Day Founding Plus** pass for sale, suggesting it is
some sort of special offer not generally available to the public. People who
hold this pass have notably more stopovers per trip than any other membership,
suggesting that whomever has one makes good use of it.

The 365 Plus
pass **is** available for sale, explaining its apparent popularity with
people who regularly need more time on their rides.

Let's look deeper into these *founding plus* members - are there any notable
geographic patterns?

```{r departing traffic by 365 plus and founding plus}
anim <- df %>% 
  filter(membership %in% c("365 Day Founding Plus", "365 Plus"), 
         id_depart != id_return) %>%  # remove round trips
  animated.map(direction = "departing",
               title = "Traffic departing from Mobi bike stations, June-July 2021     Hour: {next_state}\n365 Plus and 365 Founding Plus only")

FPS = 20; W = 960; H = 660; DTL = 1; NFRAMES = 1*24*9 # bins per hour * 24 hours * frames/hour
animate(plot = anim, fps = FPS, nframes = NFRAMES, detail=DTL, width=W, height=H)
```

for 365 plus/founding, the animation wasn't insightful. The trends were
very weak.


## All stations
All traffic departing from / arriving at stations, averaged out.

```{r departing traffic by all members}
anim <- df %>% 
  filter(id_depart != id_return) %>% 
  animated.map(direction = "departing", arrow.scale = 0.25, transition_frames = 10, state_frames = 1)

FPS = 30; W = 780; H = 660; DTL = 2; S = paste0(W,"x",H); NFRAMES = 1*24*11 # bins per hour * 24 hours * frames/hour
FPS = 30; W = 1080; H = 1080; DTL = 2; S = paste0(W,"x",H); NFRAMES = 1*24*11 # bins per hour * 24 hours * frames/hour
# animate(plot = anim, fps = FPS, nframes = NFRAMES, detail=DTL, width=W, height=H)
# anim_save("img/mobi_30fps_traffic_vectors_vp9.mp4", anim, fps = FPS, nframes = NFRAMES,
#           width=W, height=H, detail=DTL, 
#           renderer = ffmpeg_renderer(options = list(s = S, `b:v` = "1200k", f = "mp4", `c:v` = "libvpx-vp9")))
anim_save("img/mobi_30fps_traffic_vectors_vp8.mp4", anim, fps = FPS, nframes = NFRAMES,
          width=W, height=H, detail=DTL, 
          renderer = ffmpeg_renderer(options = list(s = S, `b:v` = "1200k", f = "webm", `c:v` = "libvpx")))
# anim_save("img/mobi_30fps_traffic_vectors_h264.mp4", anim, fps = FPS, nframes = NFRAMES,
#           width=W, height=H, detail=DTL,
#           renderer = ffmpeg_renderer(options = list(s = S, `b:v` = "1200k", f = "mp4", `c:v` = "libx264")))
```

The easternmost station in Stanley Park with the huge westward signal is
station 105, "Stanley Park - Totem Poles". This probably represents people
riding the seawall in the legal direction and dropping off their bike at the 
Second Beach station at the west side of the park.

```{r arriving traffic by all members}
anim <- df %>% 
  filter(id_depart != id_return) %>% 
  animated.map(direction = "arriving", arrow.scale = 0.25)

FPS = 30; W = 780; H = 660; DTL = 2; S = paste0(W,"x",H); NFRAMES = 1*24*11 # bins per hour * 24 hours * frames/hour
animate(plot = anim, fps = FPS, nframes = NFRAMES, detail=DTL, width=W, height=H)
```


Most 60-minute-limit traffic averages out to traveling in to downtown, centered
on the districts east of Granville Street. Later, we'll examine
all bike traffic and see it shows a similar pattern.

In [https://andrewluyt.github.io/divvy-bikeshare/](a previous analysis of the Divvy bikeshare) 
in Chicago, I found similar bike traffic: most traffic flowed from the outlying
districts towards a central core of high activity.


```{r look for patterns in sum by hour}
x <- df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(hour, membership) %>% 
  summarize(across(where(is.numeric), sum))
x %>% ggplot(aes(hour, trip_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, trip_distance)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, n_stopovers)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(hour, stopover_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
```

```{r look for patterns in sum by weekday}
x <- df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(weekday = wday(depart_time, label = TRUE, week_start = 1), membership) %>% 
  summarize(across(where(is.numeric), sum))
x %>% ggplot(aes(weekday, trip_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, trip_distance)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, n_stopovers)) + geom_col() + facet_wrap(~membership, ncol = 4)
x %>% ggplot(aes(weekday, stopover_minutes)) + geom_col() + facet_wrap(~membership, ncol = 4)
```

## TODO: motion vectors & volume for interesting membership types

```{r}

df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(membership, hour) %>% 
  ggplot(aes(x = hour)) +
  geom_bar() +
  facet_wrap(~membership, scales = 'fixed', ncol = 4)

df %>% 
  filter(membership %in% popular_memberships) %>% 
  group_by(membership, hour) %>% 
  summarise(across(where(is.numeric), ~mean(.x, trim = 0.05))) %>% 
  ggplot(aes(x = hour, y = n_stopovers)) +
  geom_col() +
  facet_wrap(~membership, ncol = 4) +
  labs(title = "Average stopovers per hour")
```





## TODO: separate weekday & weekend
## TODO: animate short vs long rides
## TODO: animate all individual rides as thin arrows.
might have to limit to a day, or an hour in one day.

```{r departing traffic flow 30 day pass}
anim <- df %>% 
  filter(id_depart != id_return,
         membership == "30 Day Pass") %>% 
  animated.map(direction = "departing", arrow.scale = 1,
               title.override = "30 Day Pass Traffic")

FPS = 20; W = 960; H = 660; DTL = 1; NFRAMES = 1*24*9 # bins per hour * 24 hours * frames/hour
animate(plot = anim, fps = FPS, nframes = NFRAMES, detail=DTL, width=W, height=H)
```



```{r create hourly volume map}

TMP_DF <- df %>%
  # filter(id_depart != id_return) %>%  # remove round trips
  filter(month(depart_time) == 7) %>% 
  group_by(hour, id_depart) %>%
  summarise(rides = n(), 
            lon_depart = mean(lon_depart), 
            lat_depart = mean(lat_depart))
```

## TODO: separate weekday & weekend
## TODO: combine volume and direction in one animation?

```{r animate station volume}
p <-
  TMP_DF %>%
  ggplot(aes(x = lon_depart,
             y = lat_depart,
             fill = rides,
             size = rides,
             group = id_depart)) +
  geom_sf(data = MAP, mapping = aes(), inherit.aes = FALSE, fill = 'gray20') +
  geom_sf(data = STANLEY_PARK, mapping = aes(), inherit.aes = FALSE, fill = 'gray20') +
  geom_point(pch = 21, color = "black") +
  xlim(c(-123.19, -123.050)) +
  ylim(c(49.245, 49.315)) +
  theme_map_dark +
  scale_fill_viridis(option = "plasma", begin = 0.25) +
  scale_size(range = c(2, 11)) +
  annotate("text", x = -123.1684, y = 49.3117, label = "Vancouver", color = "black", cex = 18) +
  annotate("text", x = -123.169, y = 49.312, label = "Vancouver", color = "grey60", cex = 18) +
  labs(subtitle = "Average traffic, by hour, out of each station in the Mobi network.",
       caption = "Andrew Luyt, 2021  |  Source: Mobi public data") +
  transition_states(hour, transition_length = 8, state_length = 1) +
  ggtitle("Average Mobi station use, July 2021     Hour: {next_state}") +
  shadow_wake(wake_length = 0.01,  wrap = TRUE) +
  enter_fade() +
  exit_fade()

FPS = 20; W = 960; H = 660; DTL = 1; NFRAMES = 1*24*9 # bins per hour * 24 hours * frames/hour
animate(plot = p, fps = FPS, nframes = NFRAMES, detail=DTL, width=W, height=H)
```

## TODO
color or facet by membership, etc

use a different type of map like the watercolor tiles so you can get
a cool look?


```{r EDA, echo=FALSE, message=FALSE}
# mean trip time, by membership type
df %>%
  filter(! is.na(membership)) %>%
  group_by(membership) %>%
  summarise(mean_trip_minutes = mean(trip_seconds) / 60) %>%
  arrange(desc(mean_trip_minutes)) %>%
  ggplot(aes(x=reorder(membership, mean_trip_minutes), mean_trip_minutes)) +
  geom_col() +
  coord_flip()

# mean stopovers by membership
# Look at "365 Day Founding Plus" !!!
df %>%
  filter(! is.na(membership)) %>%
  group_by(membership) %>%
  summarise(mean_stopovers = mean(n_stopovers)) %>%
  arrange(desc(mean_stopovers)) %>%
  ggplot(aes(x=reorder(membership, mean_stopovers), mean_stopovers)) +
  geom_col() +
  coord_flip()

# distribution of stopovers by membership
df %>%
  filter(! is.na(membership)) %>%
  filter(n_stopovers > 0) %>%
  ggplot(aes(x=n_stopovers)) +
  geom_density() +
  facet_wrap(~membership)

# mean trip time by departure hour
# busy time is the end of, or immediately after, the workday: 5-7pm
# morning use is WAY lower. Are people commuting one way?
df %>%
  ggplot(aes(trip_seconds/ 60)) +
  geom_histogram()

# mean trip time by return hour
# busy time is still the end of, or immediately after, the workday: 5-7pm
# Is this spike in trips a one-way commute FROM work TO home?
# Do people do this because they don't want to be sweaty at work?
df %>%
  ggplot(aes(return_hour)) +
  geom_bar()

```


```{r}
df %>% 
  mutate(day = as_date(depart_time)) %>% 
  filter(between(day, max(day) - 365, max(day) - 1)) %>% 
  group_by(day) %>%
  tally() %>% 
  ggplot(aes(day, n)) +
  geom_line() +
  labs(title = "Daily trip volume")
    
```









